---
title: "The Effect of Riparian Vegetation Density on the Nitrogen Content of Surface Water"
author: Davis Chase Dudley, Ricardo Monsalve, Jack D'Antonio
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Chunk 1 ^^^


```{r import packages}

#In Chunk 2, we're loading in the packages necessary to manipulate and process our data for further analysis.

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lubridate)

```
Chunk 2 ^^^


```{r load data sets}

#In Chunk 3, we're loading all of the raw NEON data sets that we're going to be using in our project, so that they're immediately accessible and available within the .rmd. Additionally, we're tagging each with an appropriate siteID (derived from the NEON site wherein the data was collected) and a year (derived from the year within which the data was collected). 

# (Note: Each set of data is preceded by a textual title (example: #BIGC_2019) to facilitate any necessary navigation).


#BIGC_2019

Nitrate_BIGC_2019 <-read_csv("../Data/Processed/BIGC (CA-D17)/BIGC_NEON_nitrate-surfacewater/NEON.D17.BIGC.DP1.20033.001.102.100.015.NSW_15_minute.2019-07.basic.20221201T190523Z.csv") %>% 
  mutate(siteID="BIGC", year=2019)

VegCover_BIGC_2019 <- read_csv("../Data/Processed/BIGC (CA-D17)/BIGC_NEON_cover-veg-riparian/NEON.D17.BIGC.DP1.20191.001.rip_percentComposition.2019-07.basic.20230112T002213Z.csv") %>% 
  mutate(siteID="BIGC", year=2019)


#BIGC_2021

Nitrate_BIGC_2021 <-read_csv("../Data/Processed/BIGC (CA-D17)/BIGC_NEON_nitrate-surfacewater/NEON.D17.BIGC.DP1.20033.001.102.100.015.NSW_15_minute.2021-08.basic.20221201T175835Z.csv") %>% 
  mutate(siteID="BIGC", year=2021)

VegCover_BIGC_2021 <- read_csv("../Data/Processed/BIGC (CA-D17)/BIGC_NEON_cover-veg-riparian/NEON.D17.BIGC.DP1.20191.001.rip_percentComposition.2021-08.basic.20230112T001807Z.csv") %>% 
  mutate(siteID="BIGC", year=2021)


#BLUE_2019

Nitrate_BLUE_2019 <-read_csv("../Data/Processed/BLUE (OK-D11)/BLUE_NEON_nitrate-surfacewater/NEON.D11.BLUE.DP1.20033.001.112.100.015.NSW_15_minute.2019-06.basic.20221201T045009Z.csv") %>% 
  mutate(siteID="BLUE", year=2019)

VegCover_BLUE_2019 <-read_csv("../Data/Processed/BLUE (OK-D11)/BLUE_NEON_cover-veg-riparian/NEON.D11.BLUE.DP1.20191.001.rip_percentComposition.2019-06.basic.20230112T002006Z.csv") %>% 
  mutate(siteID="BLUE", year=2019)


#BLUE_2021

Nitrate_BLUE_2021 <-read_csv("../Data/Processed/BLUE (OK-D11)/BLUE_NEON_nitrate-surfacewater/NEON.D11.BLUE.DP1.20033.001.112.100.015.NSW_15_minute.2021-06.basic.20221201T013159Z.csv") %>% 
  mutate(siteID="BLUE", year=2021)

VegCover_BLUE_2021 <-read_csv("../Data/Processed/BLUE (OK-D11)/BLUE_NEON_cover-veg-riparian/NEON.D11.BLUE.DP1.20191.001.rip_percentComposition.2021-06.basic.20230112T002305Z.csv") %>% 
  mutate(siteID="BLUE", year=2021)


#LECO_2019

Nitrate_LECO_2019 <-read_csv("../Data/Processed/LECO (TN-D07)/LECO_NEON_nitrate-surfacewater/NEON.D07.LECO.DP1.20033.001.102.100.015.NSW_15_minute.2019-09.basic.20221208T002007Z.csv") %>% 
  mutate(siteID="LECO", year=2019)

VegCover_LECO_2019 <-read_csv("../Data/Processed/LECO (TN-D07)/LECO_NEON_cover-veg-riparian/NEON.D07.LECO.DP1.20191.001.rip_percentComposition.2019-09.basic.20230112T001930Z.csv") %>% 
  mutate(siteID="LECO", year=2019)


#LECO_2021

Nitrate_LECO_2021 <-read_csv("../Data/Processed/LECO (TN-D07)/LECO_NEON_nitrate-surfacewater/NEON.D07.LECO.DP1.20033.001.112.100.015.NSW_15_minute.2021-09.basic.20221208T012753Z.csv") %>% 
  mutate(siteID="LECO", year=2021)

VegCover_LECO_2021 <-read_csv("../Data/Processed/LECO (TN-D07)/LECO_NEON_cover-veg-riparian/NEON.D07.LECO.DP1.20191.001.rip_percentComposition.2021-09.basic.20230112T002003Z.csv") %>% 
  mutate(siteID="LECO", year=2021)


```
Chunk 3 ^^^


```{r Compile Data into Single Data Frame}

#In Chunk 4, all of the extraneous data from the raw data sheets is being excised, and the data pertaining to our project is being consolidated into multiple formats in order to facilitate further analysis and manipulation. These are detailed in pre-code titles below, but generally speaking, we are creating a data frame containing all of the 'surfWaterNitrateMean' data from the Nitrate_Surfacewater NEON data sheets and a separate frame containing all of the 'canopyCoverPercent' data from the Cover_Veg_Riparian NEON data sheets (these will be used in the creation of two single-variable histograms). We're then creating mean data for each of these consolidated data frames (grouped by siteID and year) before finally left_joining these mean data to create a summarized master data frame.


#Nitrate Data Consolidation (Total, then mean surfWaterNitrateMean grouped by SiteID and year)
Nitrate_Combined <- bind_rows(Nitrate_BIGC_2019, Nitrate_BIGC_2021, Nitrate_BLUE_2019, Nitrate_BLUE_2021, Nitrate_LECO_2019, Nitrate_LECO_2021)

Mean_Nitrate_Combined <- Nitrate_Combined %>% 
  drop_na() %>% 
  group_by(siteID, year) %>% 
  summarize(mean_conc_nitrate = mean(surfWaterNitrateMean), standard_error_N = sd(surfWaterNitrateMean/sqrt(n())))


#VegCover Data Consolidation (Total, then mean canopyCoverPercent grouped by SiteID and year)
VegCover_Combined <- bind_rows(VegCover_BIGC_2019, VegCover_BIGC_2021, VegCover_BLUE_2019, VegCover_BLUE_2021, VegCover_LECO_2019, VegCover_LECO_2021)

Mean_VegCover_Combined <- VegCover_Combined %>% 
  group_by(siteID, year) %>% 
  summarize(mean_veg_cover = mean(canopyCoverPercent), standard_error_V = sd(canopyCoverPercent/sqrt(n())))


#Summary Data Frame (with corresponding mean values)
Master_Data_Frame <- left_join(Mean_Nitrate_Combined, Mean_VegCover_Combined)



#Raw 'surfWaterNitrateMean' and 'canopyCoverPercent' columns for Histogram
Total_Nitrate_Conc_Combined <- Nitrate_Combined %>% 
  drop_na() %>% 
  select(surfWaterNitrateMean)

Total_VegCover_Combined <- VegCover_Combined %>% 
  select(canopyCoverPercent)


```
Chunk 4 ^^^


```{r Histogram}

# In Chunk 5, we're using the data frames containing the compiled 'surfWaterNitrateMean' and 'canopyCoverPercent' columns to create two single-variable histograms, each representative of one of our two continuous variables (both our response and our predictor variable). These histograms will give us a better understanding of the distribution of our variable values, which could help us to draw more accurate conclusions about the relationships we're observing down the line (as well as help us to explain any potential anomalies that arise).


Total_Nitrate_Conc_Combined %>%
  ggplot(aes(x=surfWaterNitrateMean))+
  geom_histogram()

Total_VegCover_Combined %>%
  ggplot(aes(x=canopyCoverPercent))+
  geom_histogram()


```
Chunk 5 ^^^


```{r Data Frame For Graphical Analysis}

#In Chunk 6, we are creating data frames containing the summarized data necessary for visualization and analysis, specifically to facilitate the creation of figures by making all of the pertinent data immediately available in a single data frame. (Note: Much of this code is the same as that contained in Chunk 4; for the purposes of our project, it made sense to summarize our data prior to compilation, and as such, much of it already exists in a grouped and summarized format).


#Nitrate Data Consolidation (Total, then mean surfWaterNitrateMean grouped by SiteID and year)
Nitrate_Combined <- bind_rows(Nitrate_BIGC_2019, Nitrate_BIGC_2021, Nitrate_BLUE_2019, Nitrate_BLUE_2021, Nitrate_LECO_2019, Nitrate_LECO_2021)

Mean_Nitrate_Combined <- Nitrate_Combined %>% 
  drop_na() %>% 
  group_by(siteID, year) %>% 
  summarize(mean_conc_nitrate = mean(surfWaterNitrateMean), standard_error_N = sd(surfWaterNitrateMean/sqrt(n())))


#VegCover Data Consolidation (Total, then mean canopyCoverPercent grouped by SiteID and year)
VegCover_Combined <- bind_rows(VegCover_BIGC_2019, VegCover_BIGC_2021, VegCover_BLUE_2019, VegCover_BLUE_2021, VegCover_LECO_2019, VegCover_LECO_2021)

Mean_VegCover_Combined <- VegCover_Combined %>% 
  group_by(siteID, year) %>%
  summarize(mean_veg_cover = mean(canopyCoverPercent), standard_error_V = sd(canopyCoverPercent/sqrt(n())))


#Summary Data Frame (with corresponding mean values)
Master_Data_Frame <- left_join(Mean_Nitrate_Combined, Mean_VegCover_Combined) %>% 
  rename(`Mean Nitrate Concentration` = mean_conc_nitrate) %>% 
  rename(`Mean Canopy Cover Percentage` = mean_veg_cover)

```
Chunk 6 ^^^


```{r Visualization 1: Scatter Plot}

#In Chunk 7, we're visualizing our data using a scatter-plot equipped with a linear trendline in order to i) perform an analysis of our project data for interpretation and ii) to provide a standard visual of the collected data for potential readers. 

#Scatter Plot
Master_Data_Frame %>%
  ggplot(aes(x=`Mean Canopy Cover Percentage`, y=`Mean Nitrate Concentration`))+
  geom_point()+
  geom_smooth(method ="lm", color = "red")+
  theme_classic()


```
Chunk 7 ^^^

Figure 1: This scatter-plot, fit with a red trendline, shows the inverse relationship between the mean percentage of vegetative cover (mean_veg_cover along the x-axis) and the mean nitrate concentration (mean_conc_nitrate along the y-axis) across three different NEON sites, each tested twice across a two-year period. 



```{r Visualizations 2 and 3: Bar Plots}

#In Chunk 8, we're visualizing our data using two separate bar-plots. The first represents the mean nitrate concentration (mean_conc_nitrate) separated by siteID (distinguishable by color) and faceted by year. The second represents the mean vegetation percent coverage (mean_veg_cover), also separated by siteID and faceted by year. These graphs are meant to provide an easily-interpreted visual for each respective measure to i) supplement our analysis and interpretation of our project data and ii) to provide a standard visual of the collected data for potential readers.


# Mean Nitrate Concentration by Site Bar Graph (Figure 2)
Master_Data_Frame %>%
  drop_na()%>%
  group_by(siteID, year)%>%
  ggplot(aes(x=siteID, y=`Mean Nitrate Concentration`, fill = siteID))+
  geom_col(position="dodge")+
  geom_errorbar(aes(ymin = `Mean Nitrate Concentration` - standard_error_N, ymax = `Mean Nitrate Concentration` + standard_error_N), width = .2)+
  facet_wrap("year")

#Mean Vegetation Cover Percentage by Site Bar Graph (Figure 3)
Master_Data_Frame %>%
  drop_na()%>%
  group_by(siteID, year)%>%
  ggplot(aes(x=siteID, y=`Mean Canopy Cover Percentage`, fill = siteID))+
  geom_col(position="dodge")+
  geom_errorbar(aes(ymin = `Mean Canopy Cover Percentage` - standard_error_V, ymax = `Mean Canopy Cover Percentage` + standard_error_V), width = .2)+
  facet_wrap("year")


```
Chunk 8 ^^^

Figure 2: This Bar plot represents the mean concentration of nitrate in surface water across three field sites (BIGC, BLUE, and LECO) in 2019 and 2021. The fill color of each bar correlates with a respective field site, and each bar includes an error bar of one standard error. Across both years, BIGC had the lowest mean nitrate concentration, whereas BLUE had the highest mean nitrate concentration.

Figure 3: This Bar plot represents the mean vegetation cover percentage across three field sites (BIGC, BLUE, and LECO) in 2019 and 2021. The fill color of each bar correlates with a respective field site, and each bar includes an error bar of one standard error. Across both years, BIGC and LECO had nearly-identical mean vegetation cover percentages (with LECO slightly leading in both instances), whereas BLUE consistently had the lowest mean vegetation cover percentage. 



```{r Statistical Analysis}

#In Chunk 9, we're conducting a statistical analysis (specifically, a linear regression with an additional correlation value calculation) of our primary figure, the scatter-plot equipped with a trendline constructed in Chunk 7. We've done this both to determine whether or not there was a statistically-significant relationship between vegetation coverage percentage (as a proxy for density) and water nitrate concentration and to determine whether or not the conduct of our research met the assumptions for a regression test.


#Linear Model
Linear_Model <- lm(mean_conc_nitrate~mean_veg_cover, data = Master_Data_Frame)
summary(Linear_Model)
plot(Linear_Model)

#Additional Resource: Correlation Value
cor(x = Master_Data_Frame$mean_veg_cover, y = Master_Data_Frame$mean_conc_nitrate)

```
Chunk 9 ^^^

The statistical analysis of our primary figure, a scatter-plot fit with a linear trendline, tested i) whether or not there was a statistically-significant relationship between vegetation coverage percentage (as a proxy for density) and water nitrate concentration, and ii) whether or not the conduct of our research met the assumptions for a regression test. The statistical analyses determined that there is a statistically significant (but minor) negative correlation with a p-value of 0.01 (such that, as mean vegetation cover percentage increased, mean nitrate concentration decreased). The analysis also evidenced, however, that the conduct of our analysis failed to meet the assumptions of a regression test, specifically in that neither the Residuals vs. Fitted Plot relationship or the Scale-Location Plot demonstrated the 'flat-line relationship' expected of data meeting the regression line assumptions (though, interestingly enough, the Q-Q Plot was relatively sound). Here, we make note of these assumption violations and move on with our analysis. 






